Simulating High-Dimensional Multivariate Data

### 2020-12-01 010-background.Rmd
3rd draft. run spell checking and edit

The `bigsimr` `R` package provides multiple algorithms to work with high-dimesional multivariate data, but all these algorithms were originally designed to support a single task:
to generate random vectors drawn from multivariate probability distribution with given marginal distributions and dependency metrics.
Specifically, our goal is to efficiently simulate a large number, $B$, of random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components and hetereogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$, where $d$ can be very large and still be computed in practically useful times.

@@

Simulating high-dimensional, non-normal, correleated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
The RNA-seq data-generating process involves counting how often a particular messenger RNA (mRNA) is expressed in a biological sample.
RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
For human derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted.
Importantly, due to inherent biological processes, gene expression data exhibits correlation --- co-expression --- across genes [@BE07; @Schissler2018].

@@

We'll illustrate our methodology using the well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgements).
For simplicity, we only consider high expressing genes.
In turn, we begin by filtering to retain the top `r (1-myProb)*100`% highest- expressing genes (in terms of median expression) of the 20,501 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples, resulting in $d=`r d`$ genes.
This gives a massive number of pairwise dependencies among the marginals (specifically, $`r choose(d,2)`$ correlation parameters).
We further process the data to illustrate our methodology's flexible and robust simulation scheme, while better aligning with the actual data-generating process, by rounding @Li2011c's RNA-seq by Expectation Maximization (RSEM) values to counts.
Table \@ref(tab:ch010-realDataTab) displays counts for three selected high-expressing genes for the first five patients' breast tumor samples.

@@

Two nonparameteric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$. 
Define

@@

Kendall's $\tau$, on the other hand, is the difference in probabilities of concordant and discordant pairs of observations $(X_i, Y_i)$ and $(X_j, Y_j)$.
By concordance we mean that orderings have the same direction (e.g., if $X_i < X_j$, then $Y_i < Y_j$) and is determinded by the ranks of the values, not the values themselves.

@@

*Correspondence among Pearson, Spearman, $\tau$ correlations*.
This is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficent, as the marginal distributions $F_i$ are intrinstic in their calculation.
But for **bivariate normal vectors**, however, the correspondence is well-known:

@@

\noindent where $U$ is a uniform random variable in $(0,1)$, and $F^{-1}_1, F^{-1}_2$ are the inverse cdf of random variables $X_1$ and $X_2$, respectively. 
For discrete random variables, define $F^{-1}$ as in Equation \@ref(eq:inverseCDF).

@@

A Gaussian copula is the case where all marginal CDFs $F_i$ are the standard normal cdf, $\Phi$. 
This representation corresponds to a multivariate normal distribution with standard normal marginal distributions and covariance matrix ${\bf R_P}$.
But since the marginals are standardized to have unit variance, this ${\bf R_P}$ is a Pearson correlation matrix. 
If $F_{{\bf R}}$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through

@@

Sklar's Theorem [@Sklar1959; @Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) *any* random vector can be obtained via transformations involving copula functions. 
For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)$ with $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, via ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$ povides $Y_i \sim F_i, \, \forall \, i$ .

>>>>

The `bigsimr` `R` package provides multiple algorithms to work with high-dimensional multivariate data, but all these algorithms were originally designed to support a single task:
to generate random vectors drawn from multivariate probability distribution with given marginal distributions and dependency metrics.
Specifically, our goal is to efficiently simulate a large number, $B$, of random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$, where $d$ can be very large and still be computed in practically useful times.

@@

Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
The RNA-seq data-generating process involves counting how often a particular messenger RNA (mRNA) is expressed in a biological sample.
RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
For human derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted.
Importantly, due to inherent biological processes, gene expression data exhibits correlation --- co-expression --- across genes [@BE07; @Schissler2018].

@@

We'll illustrate our methodology using the well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
For simplicity, we only consider high expressing genes.
In turn, we begin by filtering to retain the top `r (1-myProb)*100`% highest- expressing genes (in terms of median expression) of the 20,501 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples, resulting in $d=`r d`$ genes.
This gives a massive number of pairwise dependencies among the marginals (specifically, $`r choose(d,2)`$ correlation parameters).
We further process the data to illustrate our methodology's flexible and robust simulation scheme, while better aligning with the actual data-generating process, by rounding @Li2011 RNA-seq by Expectation Maximization (RSEM) values to counts.
Table \@ref(tab:ch010-realDataTab) displays counts for three selected high-expressing genes for the first five patients' breast tumor samples.

@@

Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$. 
Define

@@

Kendall's $\tau$, on the other hand, is the difference in probabilities of concordant and discordant pairs of observations $(X_i, Y_i)$ and $(X_j, Y_j)$.
By concordance we mean that orderings have the same direction (e.g., if $X_i < X_j$, then $Y_i < Y_j$) and is determined by the ranks of the values, not the values themselves.

@@

*Correspondence among Pearson, Spearman, $\tau$ correlations*.
This is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions $F_i$ are intrinsic in their calculation.
But for **bivariate normal vectors**, however, the correspondence is well-known:

@@

\noindent where $U$ is a uniform random variable in $(0,1)$, and $F^{-1}_1, F^{-1}_2$ are the inverse CDF of random variables $X_1$ and $X_2$, respectively. 
For discrete random variables, define $F^{-1}$ as in Equation \@ref(eq:inverseCDF).

@@

A Gaussian copula is the case where all marginal CDFs $F_i$ are the standard normal CDF, $\Phi$. 
This representation corresponds to a multivariate normal distribution with standard normal marginal distributions and covariance matrix ${\bf R_P}$.
But since the marginals are standardized to have unit variance, this ${\bf R_P}$ is a Pearson correlation matrix. 
If $F_{{\bf R}}$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through

@@

Sklar's Theorem [@Sklar1959; @Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions.
For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)$ with $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, via ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$ povides $Y_i \sim F_i, \, \forall \, i$ .

### 2020-12-01 020-simulation-algorithm.Rmd
3rd draft. run spell checking and edit

With modern parallelized computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multicore and/or graphical-processing-unit (GPU) accelerated, algorithms --- providing the fast scaleability using readily-available hardware.

@@

To maintain scaleability (SP1), our solution is to essentially avoid this complication in Pearson matching.
Since our goal is to simulate non-normal marginals, we greatly prefer the use of rank-based measures $\rho_S$ and $\tau$ from a modeling standpoint.
Further, $\rho_S$ and $\tau$'s invariance under monotone transformation (see [Background](#background)), perserves the correlation coefficients through steps 3, 4, and 5 in the NORTA algorithm above.
This eliminates the need for computing the $\binom{d}{2}$ integrals to match exactly (nothing is for free, however, as discussed in below in Section \@ref(rand-vec-gen)).

Despite all this, if one does desire to characterize dependency using Pearson correlations, simply using the target Pearson correlation matrix as the initial conditions to our proposed algorithm will lead to approximate matching in the resultant distribution [@Song00] in many practical applications.
The quality of this approximation depends on the setting, but in practice, for high-dimensional count data we find the accuracy to be adequate.
Later, we'll study the robustness of our method to this limitation in selected [Monte Carlo evalutations]({#simulations}).

@@

1. Pre-processing for nonparameteric dependency matching.  
	(i) Convert from either ${\bf R_{Spearman}}$ or ${\bf R_{Kendall}}$ into the corresponding MVN input correlation ${\bf R_{Pearson}}$.
	(ii) Check that ${\bf R_{Pearson}}$ is semi-positive definite.  
	(iii) If not compute a close semi-positive definite correlation matrix ${\bf \widetilde{R}_{Pearson}}$.  
2. Gaussian copula construction.  
	(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_{Pearson}})$.  
	(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
3. Quantile evaluations.  
	(i) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

@@

A complication often arises at this stage: the resultant matrix may become indefinite, either through numerical error or naturally occurring in the correlation conversion.
Researchers working in multivariate computation frequently encounter such difficulties and need to find a close positive (semi-)definite matrix.
A widely-available routine for this task in `R` is called `matrix::nearPD`, though it is not suitable for high dimensions.
To overcome this issue, we've developed `bigsimr::nearPSD`, a quadractically-convergent Netwon method for finding the nearest correlation matrix, developed by @QS2006.
We hope that this routine could be useful in many applications aside from our primary goal of random vector generation.

Once the target margins and algorithm inputs are determined, steps 2 and 3 are essentially a NORTA algorithm with modern high-performance computing implementations.
Specifically, step 2i uses either an efficient multicore multivariate normal simulator (the `R` package `mvnfast`[@Fasiolo2016]) or a using Google's JAX python library NumPy for graphical-processing-unit (GPU) acceleration of the Cholesky factorization and matrix multiplication (steps 3, 4 in the NORTA algorithm in the preceeding section).

## NORTA in higher dimensions

@@

There is hope, however, as @GH02 also show that by augmenting the NORTA procedure by replacing the indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$.
This provides evidence that our nearest positive semidefinite (PSD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.

>>>>

With modern parallelized computing, steps 1, 3, 4, 5 are readily implemented as high-performance, multi-core and/or graphical-processing-unit (GPU) accelerated, algorithms --- providing the fast scaleability using readily-available hardware.

@@

To maintain scaleability (SP1), our solution is to essentially avoid this complication in Pearson matching.
Since our goal is to simulate non-normal marginals, we greatly prefer the use of rank-based measures $\rho_S$ and $\tau$ from a modeling standpoint.
Further, $\rho_S$ and $\tau$'s invariance under monotone transformation (see [Background](#background)), preserves the correlation coefficients through steps 3, 4, and 5 in the NORTA algorithm above.
This eliminates the need for computing the $\binom{d}{2}$ integrals to match exactly (nothing is for free, however, as discussed in below in Section \@ref(rand-vec-gen)).

Despite all this, if one does desire to characterize dependency using Pearson correlations, simply using the target Pearson correlation matrix as the initial conditions to our proposed algorithm will lead to approximate matching in the resultant distribution [@Song00] in many practical applications.
The quality of this approximation depends on the setting, but in practice, for high-dimensional count data we find the accuracy to be adequate.
Later, we'll study the robustness of our method to this limitation in selected [Monte Carlo evaluations]({#simulations}).

@@

1. Pre-processing for nonparametric dependency matching.  
	(i) Convert from either ${\bf R_{Spearman}}$ or ${\bf R_{Kendall}}$ into the corresponding MVN input correlation ${\bf R_{Pearson}}$.
	(ii) Check that ${\bf R_{Pearson}}$ is semi-positive definite.  
	(iii) If not compute a close semi-positive definite correlation matrix ${\bf \widetilde{R}_{Pearson}}$.  
2. Gaussian copula construction.  
	(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_{Pearson}})$.  
	(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
3. Quantile evaluations.  
	(i) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

@@

A complication often arises at this stage: the resultant matrix may become indefinite, either through numerical error or naturally occurring in the correlation conversion.
Researchers working in multivariate computation frequently encounter such difficulties and need to find a close positive (semi-)definite matrix.
A widely-available routine for this task in `R` is called `matrix::nearPD`, though it is not suitable for high dimensions.
To overcome this issue, we've developed `bigsimr::nearPSD`, a quadractically-convergent Newton method for finding the nearest correlation matrix, developed by @QS2006.
We hope that this routine could be useful in many applications aside from our primary goal of random vector generation.

Once the target margins and algorithm inputs are determined, steps 2 and 3 are essentially a NORTA algorithm with modern high-performance computing implementations.
Specifically, step 2i uses either an efficient multi-core multivariate normal simulator (the `R` package `mvnfast`[@Fasiolo2016]) or a using Google's JAX python library NumPy for graphical-processing-unit (GPU) acceleration of the Cholesky factorization and matrix multiplication (steps 3, 4 in the NORTA algorithm in the preceding section).

## A note on NORTA in higher dimensions

@@

There is hope, however, as @GH02 also show that by augmenting the NORTA procedure by replacing the indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$.
This provides evidence that our nearest positive semi-definite (PSD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.

### 2020-12-01 030-bigsimr-package.Rmd
3rd draft. run spell checking and edit

The `bigsimr` package is a high-performance implementation of the proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms) for details).
When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration, while allowing advanced users to customize and automate worflows.

This subsections below describe the basic use of `bigsimr`, by stepping through a low-dimensional (2D) simulation workflow for the often-used example data set `airquality`.
This workflow proceeds from data, to estimation, simulation configuration, random vector generation, and result vizualization.
For this low-dimensional setting, we compute using a single central processing unit (CPU) as the overhead in forking the tasks to multiple cores outways the computational gains. 
Then we transition to advanced use where we briefly describe some of the high-performance features and syntax.
The section concludes with a short description of how to use `bigsimr` on a computer clusters through slurm scheduling via the `rslurm` package.

@@

We'll demonstrate the basic use and syntax of `bigsimr` through an example workflow applied to the New York air quality data set (`airquality`) included in the R `datasets` package.
First, we load the `bigsimr` library and a few other convenient data science packages, including the syntacically-elegant `tidyverse` suite of `R` packages.

@@

Figure \@ref(fig:ch030-aq-joint-dist) visualizes the bivariate relationship between Ozone and Temperature.
We aim to simulate random two-component vectors mimicking this structure. 
The margins are not normally distributed, particulary the ozone level exhibits a strong positive skew.

@@

*Specifying marginal distributions*.
Based on the estimated densities in Figure \@ref(fig:ch030-aq-joint-dist), we'll assume `Temp` is normally ditsributed and `Ozone` is log-normally distributed since its values are positive and skewed.
We'll use the well-known, unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the lognormal parameters:

@@

Notice that we use the *quantile* function for the marginals, as that is how the marginal distributions $F_i$ enter into the `bigsimr::rvec` algorthm.
This implementation strategy supports all `R` `base` probability distributions.
And allows flexible extensions using other `R` packages that adhere to conventions, such as `extraDistr`.
Further, by using `alist`, users can specify their own custom distributions (see below in [creating custom margins](#custom-margins)).

@@

*Specfying correlation*.
As mentioned, the user must decide how to describe correlation, based on the particulars of the problem.
For non-normal data and for improved simulation accuracy in our scheme, we advocate the use of rank-based correlations Spearman's $\rho_S$ and Kendall's $\tau$.
But we also support approximate Pearson correlation coefficient matching, while cautioning the user to check the performance for their parameteric multivariate model (see [Monte Carlo evaluations](#simulations) for evaluation strategies and guidance).
To aid in correlation specification, and estimation in general, we provide a high-performance function `bigsimr::cor_fast` which estimates Pearson, Spearman, or Kendall correlation using the fastest methods available.
(Anyone who has tried estimating Kendall's $\tau$ using `stats::cor` can attest that the routine does not scale to even moderate dimensions).
Notably, these estimation methods are the standard approaches, not designed specifically designed for high-dimensional correlation estimation (see [Conclusion and Discussion]({#discussion) for more on this).

@@

*Checking the theoretical correlation bounds*
As discussed in Section \@ref(background), given a pair of marignal distributions the possible correlations are not free to vary between $[-1, 1]$. 
To ensure that the simulation is not configured to impossible settings, we provide the `bigsimr::cor_bounds` fuction provides MC estimated theoretical lower and upper bounds (using the Generate, Sort, and Correlate algorithm of @DH2011).

@@

```{r ch030-plot-sim, fig.cap="Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels. The simulated points mimick the observed data with respect to both the marginal characteristics and bivariate association."}
p1 <- df_sim %>%
  ggplot(aes(Temp, Ozone)) +
  geom_density_2d_filled() +
  theme(legend.position = "none") +
  labs(x = "Simulated Temperature", y = "Simulated Ozone") +
  scale_y_continuous(limits = c(0, max(df$Ozone))) +
  scale_x_continuous(limits = range(df$Temp))

@@

*Creating a custom marginal distribution*.
Because `bigsimr` uses an `alist` to store the margins, any probability distribution with a well-defined inverse CDF can be used including custom marginal distributions not provided in base `R`. 
Therefore, to specify a marginal distribution absent from an existing packge, the user needs to provide a closed-form expression to form the corresponding quantile function.

>>>>

The `bigsimr` package is a high-performance implementation of the proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms) for details).
When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration, while allowing advanced users to customize and automate workflows.

This subsections below describe the basic use of `bigsimr`, by stepping through a low-dimensional (2D) simulation workflow for the often-used example data set `airquality`.
This workflow proceeds from data, to estimation, simulation configuration, random vector generation, and result visualization.
For this low-dimensional setting, we compute using a single central processing unit (CPU) as the overhead in forking the tasks to multiple cores outweighs the computational gains. 
Then we transition to advanced use where we briefly describe some of the high-performance features and syntax.
The section concludes with a short description of how to use `bigsimr` on a computer clusters through slurm scheduling via the `rslurm` package.

@@

We'll demonstrate the basic use and syntax of `bigsimr` through an example workflow applied to the New York air quality data set (`airquality`) included in the R `datasets` package.
First, we load the `bigsimr` library and a few other convenient data science packages, including the syntactically-elegant `tidyverse` suite of `R` packages.

@@

Figure \@ref(fig:ch030-aq-joint-dist) visualizes the bivariate relationship between Ozone and Temperature.
We aim to simulate random two-component vectors mimicking this structure. 
The margins are not normally distributed, particularly the ozone level exhibits a strong positive skew.

@@

*Specifying marginal distributions*.
Based on the estimated densities in Figure \@ref(fig:ch030-aq-joint-dist), we'll assume `Temp` is normally distributed and `Ozone` is log-normally distributed since its values are positive and skewed.
We'll use the well-known, unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the log-normal parameters:

@@

Notice that we use the *quantile* function for the marginals, as that is how the marginal distributions $F_i$ enter into the `bigsimr::rvec` algorithm.
This implementation strategy supports all `R` `base` probability distributions.
And allows flexible extensions using other `R` packages that adhere to conventions, such as `extraDistr`.
Further, by using `alist`, users can specify their own custom distributions (see below in [creating custom margins](#custom-margins)).

@@

*Specifying correlation*.
As mentioned, the user must decide how to describe correlation, based on the particulars of the problem.
For non-normal data and for improved simulation accuracy in our scheme, we advocate the use of rank-based correlations Spearman's $\rho_S$ and Kendall's $\tau$.
But we also support approximate Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) for evaluation strategies and guidance).
To aid in correlation specification, and estimation in general, we provide a high-performance function `bigsimr::cor_fast` which estimates Pearson, Spearman, or Kendall correlation using the fastest methods available.
(Anyone who has tried estimating Kendall's $\tau$ using `stats::cor` can attest that the routine does not scale to even moderate dimensions).
Notably, these estimation methods are the standard approaches, not designed specifically designed for high-dimensional correlation estimation (see [Conclusion and Discussion]({#discussion) for more on this).

@@

*Checking the theoretical correlation bounds*
As discussed in Section \@ref(background), given a pair of marginal distributions the possible correlations are not free to vary between $[-1, 1]$. 
To ensure that the simulation is not configured to impossible settings, we provide the `bigsimr::cor_bounds` function provides MC estimated theoretical lower and upper bounds (using the Generate, Sort, and Correlate algorithm of @DH2011).

@@

```{r ch030-plot-sim, fig.cap="Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels. The simulated points mimic the observed data with respect to both the marginal characteristics and bivariate association."}
p1 <- df_sim %>%
  ggplot(aes(Temp, Ozone)) +
  geom_density_2d_filled() +
  theme(legend.position = "none") +
  labs(x = "Simulated Temperature", y = "Simulated Ozone") +
  scale_y_continuous(limits = c(0, max(df$Ozone))) +
  scale_x_continuous(limits = range(df$Temp))

@@

*Creating a custom marginal distribution*.
Because `bigsimr` uses an `alist` to store the margins, any probability distribution with a well-defined inverse CDF can be used including custom marginal distributions not provided in base `R`. 
Therefore, to specify a marginal distribution absent from an existing package, the user needs to provide a closed-form expression to form the corresponding quantile function.

### 2020-12-01 040-monte-carlo-evaluation.Rmd
3rd draft. run spell checking and edit

Figure \@ref(fig:ch040-bPlot) displays the aggregated bivariate simulation results with a detailed summarization.
Table \@ref(tab:ch040-biMAEtable) contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
Taken together, the studies show our methodology is generally accuracy across the entire range of possible correlation values for the rank-based dependency measures, at least in these limited simulation settings for the rank-based correlations.
For the two non-normal bivariate marginals, the Pearson correlation matching is approximate.
For discrete margins, matching the dependency measures was somewhat less accurate, even for the rank-based metrics, and particularly inaccurate near the lower bound of Pearson correlations.

```{r ch040-bPlot, cache=F, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.asp = 1.30, fig.weight = 4, fig.align='center', fig.cap = "Bivariate simulations match specified correlations. The horizontal axis plots the specified target correlations across the entire range of possible correlations for each bivariate margin. Normal margins are plotted in dark blue, gamma in medium blue, and negative binomial in light blue. As the number of simulated vectors $B$ increases from left to right, the variation in estimated correlations (vertical axis) decreases. The dashed line indicates equality between the specified and estimated correlations. Only the normal margins match the Pearson correlations (top row) exactly across the entire range of correlations. The two non-normal bivariate random vectors experience attentuation (downward bias) for extreme Pearson correlations (due to limitations in our algorithm) and more restriction (due to the Frechet bounds). The rank-based correlations (bottom two rows) are matched exactly for all margins and obtain the full range of possible dependencies."}
# https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
# New facet label names
repsLabs <- paste0("B=", c("1,000", "10,000", "100,000") )
names(repsLabs) <- c(1000, 10000, 100000)
typeLabs <- c( 'Pearson', 'Spearman', 'Kendall' )
names(typeLabs) <- c( 'Pearson', 'Spearman', 'Kendall' )

>>>>

Figure \@ref(fig:ch040-bPlot) displays the aggregated bivariate simulation results.
Table \@ref(tab:ch040-biMAEtable) contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.
Taken together, the studies show our methodology is generally accuracy across the entire range of possible correlation values for the rank-based dependency measures, at least in these limited simulation settings for the rank-based correlations.
For the two non-normal bivariate marginals, the Pearson correlation matching is approximate.
For discrete margins, matching the dependency measures was somewhat less accurate, even for the rank-based metrics, and particularly inaccurate near the lower bound of Pearson correlations.

```{r ch040-bPlot, cache=F, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.asp = 1.30, fig.weight = 4, fig.align='center', fig.cap = "Bivariate simulations match specified correlations. The horizontal axis plots the specified target correlations across the entire range of possible correlations for each bivariate margin. Normal margins are plotted in dark blue, gamma in medium blue, and negative binomial in light blue. As the number of simulated vectors $B$ increases from left to right, the variation in estimated correlations (vertical axis) decreases. The dashed line indicates equality between the specified and estimated correlations. Only the normal margins match the Pearson correlations (top row) exactly across the entire range of correlations. The two non-normal bivariate random vectors experience attenuation (downward bias) for extreme Pearson correlations (due to limitations in our algorithm) and more restriction (due to the Frechet bounds). The rank-based correlations (bottom two rows) are matched exactly for all margins and obtain the full range of possible dependencies."}
# https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
# New facet label names
repsLabs <- paste0("B=", c("1,000", "10,000", "100,000") )
names(repsLabs) <- c(1000, 10000, 100000)
typeLabs <- c( 'Pearson', 'Spearman', 'Kendall' )
names(typeLabs) <- c( 'Pearson', 'Spearman', 'Kendall' )

### 2020-12-01 050-examples.Rmd
3rd draft. run spell checking and edit

This section demonstrates how to simulate multivariate data using `bigsimr`, aiming to replicate the structure of high-dimensional dependent count data.
In an illustration of our proposed methodology applied to real data, we seek to simulate RNA-sequencing data by producing simulated random vectors with assumed marginal distributions with estimated parameters that mimick the observed data and its generating process.
We seek scaleable realistic multivariate simulation to enable Monte Carlo (MC) methods for thes data.
Modeling RNA-seq using multivariate probabilities distributions is natural as inter-gene correlation is an inherent part of biological processes.
Yet many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction.
See @Efron2012 for a detailed discussion with related methods and see @Wu2012b, @Schissler2018, @Schissler2019 for applied examples.
The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicated an estimated parametric structure, MC probability estimation, and MC evaluation of correlation estimation efficiency.

@@

Next, we estimate the marginal parameters. 
We use method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
The marginal distributions are from the same probability family (NB) yet are hetereogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
The functions below help achieve this estimation and specifying the inputs for use in `bigsimr::rvec`.

@@

Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes in Table \@ref(tab:ch010-realDataTab).
Simulated counts roughly mimick the observed data but with a smoothier appearance due to the assumed parameter form and with less extreme points then the observed data (c.f. Figure \@ref(fig:ch010-realDataFig).)

@@

where $\chi_i$ is the sample space for the $i^{th}$ component of the random vector $\bf{Y}$.
Compact representations with convenient computational forms are rare for high-dimensional constructions, especially with hetereogeneous, correlated marginal distributions (or margins of mixed data types).
Given a large number simulated vectors as produced above, estimated probabilities are readily given by counting the proportion of simulated vectors meeting the desired condition.
In our motivating application, one may ask what is the probability that all genes expressed greater than a certain threshold value ${ \bf y}_0$.

>>>>

This section demonstrates how to simulate multivariate data using `bigsimr`, aiming to replicate the structure of high-dimensional dependent count data.
In an illustration of our proposed methodology applied to real data, we seek to simulate RNA-sequencing data by producing simulated random vectors with assumed marginal distributions with estimated parameters that mimic the observed data and its generating process.
We seek scaleable realistic multivariate simulation to enable Monte Carlo (MC) methods for these data.
Modeling RNA-seq using multivariate probabilities distributions is natural as inter-gene correlation is an inherent part of biological processes.
Yet many models do not account for this, leading to major disruptions to the operating characteristics of statistical estimation, testing, and prediction.
See @Efron2012 for a detailed discussion with related methods and see @Wu2012b, @Schissler2018, @Schissler2019 for applied examples.
The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicated an estimated parametric structure, MC probability estimation, and MC evaluation of correlation estimation efficiency.

@@

Next, we estimate the marginal parameters. 
We use method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
The marginal distributions are from the same probability family (NB) yet are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
The functions below help achieve this estimation and specifying the inputs for use in `bigsimr::rvec`.

@@

Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes in Table \@ref(tab:ch010-realDataTab).
Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parameter form and with less extreme points then the observed data (c.f. Figure \@ref(fig:ch010-realDataFig).)

@@

where $\chi_i$ is the sample space for the $i^{th}$ component of the random vector $\bf{Y}$.
Compact representations with convenient computational forms are rare for high-dimensional constructions, especially with heterogeneous, correlated marginal distributions (or margins of mixed data types).
Given a large number simulated vectors as produced above, estimated probabilities are readily given by counting the proportion of simulated vectors meeting the desired condition.
In our motivating application, one may ask what is the probability that all genes expressed greater than a certain threshold value ${ \bf y}_0$.

### 2020-12-01 060-conclusion-discussion.Rmd
3rd draft. run spell checking and edit

We've introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a high-performance `R` package called [`bigsimr`](https://schisslergroup.github.io/bigsimr/).
The random vector generation method is largely inspired by NORTA (@Cario1997) and Gaussian copula-based approaches (@MB13, @BF17, @Xia17).
The major contributions of this work are high-dimensional scalability, flexible modeling of dependency, and an high-performance implementation with broad potential data analytic applications for modern, big-data statistical computing.

It is customary to compare new tools and algorthms directly to existing competing methods and software.
In this study, however, we only employ our proposed methodology, since our previous work has show that existing tools are simply not designed or feasible to meet our high-dimensional goal (see @Li2019gpu for evalutions of the `R` `copula`package and others). 
For the bivariate simulations, existing packages such as [`nortaRA`](https://github.com/cran/NORTARA/blob/master/inst/doc/NORTARA.R) work well to match Pearson correlations exactly.

>>>>

We've introduced a general-purpose high-dimensional multivariate simulation algorithm and provide a high-performance `R` package called [`bigsimr`](https://schisslergroup.github.io/bigsimr/).
The random vector generation method is largely inspired by NORTA (@Cario1997) and Gaussian copula-based approaches (@MB13, @BF17, @Xia17).
The major contributions of this work are high-dimensional scaleability, flexible modeling of dependency, and an high-performance implementation with broad potential data analytic applications for modern, big-data statistical computing.

It is customary to compare new tools and algorithms directly to existing competing methods and software.
In this study, however, we only employ our proposed methodology, since our previous work has show that existing tools are simply not designed or feasible to meet our high-dimensional goal (see @Li2019gpu for evaluations of the `R` `copula`package and others). 
For the bivariate simulations, existing packages such as [`nortaRA`](https://github.com/cran/NORTARA/blob/master/inst/doc/NORTARA.R) work well to match Pearson correlations exactly.

### 2020-12-01 070-supplementary-materials.Rmd
edit overall. Cut RNA-seq MSE sizes. Revise RNA-seq example heavily.
3rd draft. Finalizing pdf output and small issues
Draft, revise NORTA in higher d, cite GH02
3rd draft. run spell checking and edit

# Supplementary Materials {#misc}

We provide an open-source implementation of our methology as the `bigsimr` R package, hosted on github, https://schisslergroup.github.io/bigsimr/.

# Acknowledgement(s)

@@

# Disclosure statement

@@

# Funding

>>>>

# Supplementary Materials {.unnumbered #misc}

We provide an open-source implementation of our methodology as the `bigsimr` R package, hosted on github, https://schisslergroup.github.io/bigsimr/.

```{r, out.width='5%', fig.align='center'}
knitr::include_graphics('images/hex-bigsimr.png')
```

# Acknowledgment(s) {.unnumbered #acknowledgments}

@@

# Disclosure statement {.unnumbered #coi}

@@

# Funding {.unnumbered #funding}
